### 2019-09-23 reports/article.Rmd
Use online csl
ready to submit

---
bibliography: P:/library.bib
output: 
  bookdown::word_document2:
    toc: false
    df_print: kable
    reference_docx: mall.dotx
csl: jama.csl
---

@@

*Correspondence:* Erik Bulow, SHPR, Registercentrm Vastra Gotaland, 415 43 Gothenburg, Sweden.
Tel: +46 7008234 28. E-mail: erik.bulow@rccvast.se

@@

Comorbidity indices (such as proposed by Elixhauser et al.[@Elixhauser1998] or Charlson et al.[@Charlson1987]) are common components of joint replacement studies. It has been showed however that they add little to the understanding of health-related quality of life [@Greene2015a], occurrence of re-operations [@Gordon2013] or mortality [@Bulow2017]. Beside considering comorbidity indices and joint replacement, these papers have yet another common theme, they are predictive studies, and their assessment strictly applies to predictions. Prediction and estimation are often used interchangeably, although a clear distinction should be made [@Shmueli2011]. Predictive studies or risk calculators aim to predict the outcome for a specific patient (or groups of patients with similar characteristics). Baseline comorbidity indices seemingly adds little in this case. Etiological studies of association between exposures and outcomes on the other hand, could still benefit from considering comorbidity on a population level. Such studies have showed that comorbidity indices are associated with survival [@Glassou2017], as well as medical expenses [@Penna2019], for patients on average.

@@

These patients were linked to the Swedish National Patient Register (NPR) [@Cnudde2016] which provided ICD-10 codes recorded at hospital visits for each patietn during one year before surgery. Individual ICD-10 codes were first identified as different comorbidities, then combined by the Elixhauser Comorbidity classification. The number of identified comorbidities were summed to an un-weighted index. The window for inclusion ICD-10 codes was set from 366 to 1 day prior to surgery. Patients with no recorded pre-opetative hospital visit were assumed to have no comorbidity, thus index = 0. For further details on the patient data please see [@Bulow2017].

@@

An alternative to $E[T]$ is the $\tau$-restricted mean survival time (RMST). Here, we do not aim to follow all patients to their death, but only to a pre-defined length of time, $\tau \leq T$.

RMST is estimated as $\mu(\tau)= \int_0^\tau S(t)dt$, and is the average number of years (or a period of any other time unit) survived before time $\tau$. A natural estimator for $\mu(τ)$ is $\hat \mu(\tau) = \int _0^\tau \hat S (t) dt$  where we the survival function $S$ is estimated by $\hat S$, the Kaplan-Meier curve.

@@

Instead of looking at the life expectancy up to a certain time point, we could also estimate the time lost up to that timepoint, the Restricted mean time lost ($RMTL(\tau) = \tau-RMST(\tau)$. If $\tau$ is known and fixed, then $RMTL(\tau)$ is normally distributed with variance $$\sigma_{RMLT(\tau)}^2 = \sigma_{RMST(\tau)}^2 = 2 \int_0^tS(t)dt - [\int_0^ttS(t)dt ]^2$$ [@Royston2013].

$RMST$ and $RMTL$ are asymptotically normally distributed [@Andersen2012], so standard statistical routines can be used for confidence interval building and statistical inference.

@@

There were `r N` patients included in the study (Figure \@ref(fig:flowchart)). `r p[1]` had no comorbidities, `r p[2]` had one, `r p[3]` had two, `r p[4]` three and `r p[5]` had four or more (Table \@ref(tab:table1)).
From the onset of the follow-up we saw a clear association between survival and the Elixhauser comorbidity index (Figure \@ref(fig:rmst)). The expected life time lost (RMTL) increased with comorbidity (Table \@ref(tab:tableres)). Differences were statistically significant. Although, at the beginning of the follow-up this difference was small. The time lost in the first 30 days after  hip replacement did not exceed `r ceiling(max(rmtl30d$rmtl))` hours. This differnece is most likely clinically irrelevant. At 10 years however, patients with no pre-operative comorbidity (Elixhauser = 0) had lost less than `r ceiling(min(rmtl10y$rmtl))` years of their expected life time, while patients with at least four comorbidities lost more than `r floor(max(rmtl10y$rmtl))` years. This differnece is clinically relevant.

The restricted mean time lost was between factor `r min(rr$ll)` and factor `r max(rr$ul)` higher in groups with comorbidities than in patients with Elixhauser index = 0. This was less notable with passing time (Figure \@ref(fig:rmtl)).

@@

Neither a clinician nor a survival curve can accurately predict how long a patient will live. We have also previsouly showed that ranking of individual survival times based on comorbidity, such as the Elixhauser index, have low precision [@Bulow2017; @Bulow2019]. We have seen here however that there is a clear separation on group level between patients with different Elixhauser scores. Thus, for health administrative purposes there might be added value in considering comorbidities, although less so for individual patient predictions.

There is a known increase in short-term mortality after hip replacement [@Pedersen2011; @Cnudde2018b], which is likely influenced by patient co-morbidity. We also saw that the relative RMTL was higher in the beginning of the follow-up period, compared to the end (Figure \@ref(fig:rmtl)), but that the absolute values of RMST decreased with time for all comorbidity groups (Figure \@ref(fig:rmst)). The 90-day RMTL for patients with at least four comorbidities was only `r round(max(rmtl90d$rmtl))` day, approximatively `r max(rr$ul)` times longer than for patients without comorbidity. This relative difference decreased with time. At 10-years follow up, patients with at least four comorbidities lost `r  max(rmtl_rr[rmtl_rr$years == 10, "rr"])` times as mush life time, compared to the group without comorbidity. This convergence of relative RMTL coincides with previsoulsy reported time-decreasing predictive power of comorbidity on mortality [@Bulow2017].

If we follow the dichotomy by Shmueli et al. [@Shmueli2011] we could conclude that predictive studies/risk calculators gain litle from considering the Elixhauser comorbidity index. However, for studies that aim to estimate effects of different treatment options, comorbidity indices can be important confounders and are important for case-mix adjustments [@Inacio2015a]. Health-care administrators for example might gain knowledge from considering comorbidities to better estimate the future need of revisions. The revision rate of hip replacements is low [@Bayliss2017] but patient survival increase over time [@Schwartsmann2015]. It is thererefore important to estimate the long-term survival for patietns at risk, wherefor comorbidities are relevant.

@@

@@

```{r table1}
load("../cache/table1.RData")
tableone::kableone(table1, caption = "Demography for the 150,367 patients with total hip arthroplasty.")
```

```{r tableres}
load("../cache/table_rmst_rmtl.RData")
knitr::kable(table_rmst_rmtl, caption = "Restricted mean time survival (RMST) and restricted mean time lost (RMTL) for the 150,367 patients with total hip arthroplasty.")
```

@@

```{r rmst, fig.cap = "Restricted mean time lost during 10 years follow-up stratified by the Elixhauser comorbidity index."}
knitr::include_graphics("../graphs/rmst.png")
```

```{r rmtl, fig.cap = "Relative change in restricted mean time lost with increasing Elixhauser comorbidity index with comorbidity free patients as reference."}
knitr::include_graphics("../graphs/rmtl.png")
```

>>>>

---
bibliography: P:/library.bib
output: 
  bookdown::word_document2:
    toc: false
    df_print: kable
    reference_docx: mall.dotx
csl: http://www.zotero.org/styles/jama
---

@@

*Correspondence:* Erik Bulow, SHPR, Registercentrum Vastra Gotaland, 415 43 Gothenburg, Sweden.
Tel: +46 7008234 28. E-mail: erik.bulow@rccvast.se

@@

Comorbidity indices (such as proposed by Elixhauser et al.[@Elixhauser1998] or Charlson et al.[@Charlson1987]) are common components of joint replacement studies. It has been showed however that they add little to the understanding of health-related quality of life [@Greene2015a], occurrence of re-operations [@Gordon2013] or mortality [@Bulow2017]. Beside considering comorbidity indices and joint replacement, these papers have yet another common theme, they are predictive studies, and their assessment strictly applies to predictions. Prediction and estimation are often used interchangeably, although a clear distinction should be made [@Shmueli2011]. Predictive studies or risk calculators aim to predict the outcome for a specific patient (or groups of patients with similar characteristics). Baseline comorbidity indices seemingly adds little in this case. Etiological studies of association between exposures and outcomes on the other hand, could still benefit from considering comorbidity on a population level. Such studies have showed that comorbidity indices are associated with survival [@Glassou2017], as well as with medical expenses [@Penna2019], for patients on average.

@@

These patients were linked to the Swedish National Patient Register (NPR) using the unique Swedish identity numbers [@Cnudde2016]. NPR provided ICD-10 codes recorded at hospital visits for each patient during one year before surgery. Individual ICD-10 codes were first identified as different comorbidities by the Elixhauser comorbidity classification [@Quan2005]. The number of identified comorbidities were summed to an un-weighted index. The window for inclusion ICD-10 codes was set from 366 to 1 day prior to surgery. Patients with no recorded pre-operative hospital visit were assumed to have no comorbidity, thus Elixhauser = 0. For further details on the patient data please see [@Bulow2017].

@@

An alternative to $E[T]$ is the $\tau$-restricted mean survival time (RMST). Here, we do not aim to follow all patients to their death, $T$, but only to a pre-defined length of time, $\tau \leq T$.

RMST $= \mu(\tau)= \int_0^\tau S(t)dt$ where we estimate the survival $S$ by Kaplan-Meier. It is the average number of years (or a period of any other time unit) survived before time $\tau$.

@@

Instead of looking at the life expectancy up to a certain time point, we could also estimate the time lost up to that time-point, the restricted mean time lost ($RMTL(\tau) = \tau-RMST(\tau)$). If $\tau$ is known and fixed, both RMST and RMTL are asymptotically normally distributed [@Andersen2012] with variance [@Royston2013] $\sigma^2 = 2 \int_0^\tau S(t)dt - [\int_0^\tau tS(t)dt ]^2$. Standard routines can therefore be used for confidence intervals and statistical inference.

@@

There were `r N` patients included in the study (Figure \@ref(fig:flowchart)). `r p[1]` had no comorbidities, `r p[2]` had one, `r p[3]` had two, `r p[4]` three and `r p[5]` had four or more (Table \@ref(tab:table1)).
From the onset of the follow-up we saw a clear association between survival and the Elixhauser comorbidity index (Figure \@ref(fig:rmst)). The expected life time lost (RMTL) increased with comorbidity (Table \@ref(tab:tableres)). Differences were statistically significant. Although, at the beginning of the follow-up this difference was small. The time lost in the first 30 days after  hip replacement did not exceed `r ceiling(max(rmtl30d$rmtl))` hours. This difference is clinically irrelevant. At 10 years however, patients with no pre-operative comorbidity (Elixhauser = 0) had lost less than `r ceiling(min(rmtl10y$rmtl))` year of life time, while patients with at least four comorbidities lost more than `r floor(max(rmtl10y$rmtl))` years. This difference is clinically relevant.

RMTL was between factor `r min(rr$ll)` and factor `r max(rr$ul)` higher in groups with comorbidities compared to patients without. The difference was less notable with passing time (Figure \@ref(fig:rmtl)).

@@

Neither a clinician nor a survival curve can accurately predict how long a patient will live. We have also previously showed that ranking of individual survival times based on comorbidity, such as the Elixhauser index, have low precision [@Bulow2017; @Bulow2019]. We have seen here however that there is a clear separation on group level between patients with different Elixhauser scores. Thus, for health administrative purposes there might be added value in considering comorbidities, although less so for individual patient predictions.

There is a known increase in short-term mortality after hip replacement [@Pedersen2011; @Cnudde2018b], which is likely influenced by patient co-morbidity. We also saw that the relative RMTL was higher in the beginning of the follow-up period, compared to the end (Figure \@ref(fig:rmtl)), but that the absolute values of RMST decreased with time for all comorbidity groups (Figure \@ref(fig:rmst)). The 90-day RMTL for patients with at least four comorbidities was only `r round(max(rmtl90d$rmtl))` day, approximately `r max(rr$ul)` times longer than for patients without comorbidity. This relative difference decreased with time. At 10-years follow up, patients with at least four comorbidities lost `r  max(rmtl_rr[rmtl_rr$years == 10, "rr"])` times as mush life time, compared to the group without comorbidity. This convergence of relative RMTL coincides with previously reported time-decreasing predictive power of comorbidity on mortality [@Bulow2017].

If we follow the dichotomy by Shmueli et al. [@Shmueli2011] we could conclude that predictive studies/risk calculators gain little from considering the Elixhauser comorbidity index. However, for studies that aim to estimate effects of different treatment options, comorbidity indices can be important confounders and are important for case-mix adjustments [@Inacio2015a]. Health-care administrators for example might gain knowledge from considering comorbidities to better estimate the future need of revisions. The revision rate of hip replacements is low [@Bayliss2017] but patient survival increase over time [@Schwartsmann2015]. It is therefore important to estimate the long-term survival for patients at risk, wherefore comorbidities are relevant.

It could be argued that the absence of covariate adjustment is a limitation to the study. It was seen in table \@ref(tab:table1) that age and sex are also associated with death and we have showed earlier that those variables are better predictors of long-term survival [@Bulow2017; @Bulow2019]. Theoretical methods for such adjustment have been suggested in the litterature [@Tian2014; @Conner2018] but software implementations are limited and sub-optimal. We therefore focused on crude estimates, which we thoughgt were most relevant to this study.

@@

# Table and figure captions

@@

```{r table1}
load("../cache/table1.RData")
tableone::kableone(
  table1, 
  caption = sprintf("Demography for the %s patients with total hip arthroplasty.", N)
)
```

```{r tableres}
load("../cache/table_rmst_rmtl.RData")
knitr::kable(
  table_rmst_rmtl, 
  caption = sprintf("Restricted mean time survival (RMST) and restricted mean time lost (RMTL) for the %s patients with total hip arthroplasty. RMST + RMTL should equal the whole period in each column. RMST decrease by increased comorbidity.", N)
  )
```

@@

```{r rmst, fig.cap = "Restricted mean survival time (RMST) during 10 years follow-up stratified by the Elixhauser comorbidity index."}
knitr::include_graphics("../graphs/rmst.png")
```

```{r rmtl, fig.cap = "Relative change in restricted mean time lost (RMLT) with increasing Elixhauser comorbidity index with comorbidity free patients as reference."}
knitr::include_graphics("../graphs/rmtl.png")
```
